---
title: "Google Flu Example SEIRS Model"
author: "Grant Brown"
date: "07/11/2014"
output: html_document
---

Part 0: Introduction
============================

Google Flu Trends provides estimates of influenza like activity for the United States, among other places. 
More information is available at <http://www.google.org/flutrends/>. Most importantly for this writing, Google Flu trends provides an easily 
accessible data set to test epidmic models using the spatialSEIR R library.  

The spatialSEIR R library is a rapidly developing tool for fitting spatial SEIRS models, as well as several important spatial and non-spatial special cases:

* spatial and non-spatial SEIR models
* spatial and non-spatial serial SEIR models (re-start process between each epidemic)
* spatial and non-spatial fixed rate immunity loss SEIRS models

Here we're going to fit a general, non-spatial, SEIRS model using data from Google Flu.

```{r tidy=FALSE}
library(spatialSEIR)
```


Part 1: Data Processing
=========================

The first step is to read in and process the data. For this tutorial, we're going to consider just states which are part of Health and Human Services region 5. These are:

* Illinois: population 12,830,632 (2010)
* Indiana: population 6,483,802 (2010)
* Ohio: population 11,536,504 (2010)  
* Michigan: population 9,883,640 (2010)
* Minnesota: population 5,303,925 (2010)
* Wisconsin: population 5,686,986 (2010)

Population estimates were obtained from the Census Bureau: <http://quickfacts.census.gov/qfd/states/>.
Let's consider data between 2007 and 2011.

```{r, tidy=FALSE}
fluDat = read.table("http://www.google.org/flutrends/us/data.txt", 
                    sep = ",", skip = 11, head=TRUE)
dat.keep = ((as.Date(fluDat$Date) < as.Date("2011-01-01")) & 
              (as.Date(fluDat$Date) >= as.Date("2007-01-01")))
fluDat.sub = fluDat[dat.keep,]
region5 = fluDat.sub$HHS.Region.5..IL..IN..MI..MN..OH..WI.
```

Let's take a look at the raw data. 

```{r, echo=TRUE, tidy=FALSE}
par(xaxt="n") # Suppress plotting the X axis
plot(region5, type = "l", 
     main = "ILI Per 100k Physician Visits\n HHS Region 5", 
     xlab = "Year", ylab = "#Visits/100k")
abline(h=seq(0, 10000,1000), lty = 2, col = "lightgrey")
lines(region5, col = "blue", lwd = 1.5)
par(xaxt="s") # Enable plotting of the X axis
years = as.numeric(format(as.Date(fluDat.sub$Date), "%Y"))
yearIdx = which(!duplicated(years))
# We're going to make several axes, so put it in a function
xAxis = function()
{
  par(xaxt="s")  
  axis(side=1,at=yearIdx, labels = years[yearIdx])
}
xAxis()
```

This data is presented on the "proportion of physiscian visits" scale, but we're interested in the number of new cases in the *population*. The formal approach to this dilemma is to formulate an explicit *data model* linking the observed data and the rest of the *process model*. At the time of this writing, however, libspatialSEIR does not support any but the simplest (identity) data model. Therefore we need to come up with a reasonable estimate of the number of new infectious cases. 

If we assume that the proportion of physician visits reflecting ILI is representative of the proportion of the population becoming newly infected (a questionable assumption that is nevertheless reflected in many more formal data models), we can use the population size to estimate the newly infectious counts. 


```{r, tidy=FALSE}

TotalPop =(12830632
          + 6483802
          + 11536504 
          + 9883640
          + 5303925 
          + 5686986)
FluCountEst = (region5/100000)*TotalPop

```


Part 2: SEIRS Preparation
==========================

Now that we have data on a reasonable scale for a SEIRS model, we need to make some decisions about what the infection and reinfection linear models looks like. 

From this point onward things are going to be formatted for libspatialSEIR. There remains a good deal of API cleanup and model component abstraction to be done, so be warned that the following structure is not particularly streamlined. 

2a: Infection Structure
-------------------------
The infection intensity is a function of both the infectivity of the current influenza strain and the aggregate effect of population behaviors. We could introduce variabels reflecting the introduction of public health campaigns, measures of public awareness of particular influenza outbreaks, or any number of other things (so long as we're careful not to introduce collinearity problems). 

Here we'll keep things simple, and include (in addition to an intercept) a linear term for time to capture average increases or decreases in infection intensity along with a trigonometric basis with a period of one year to capture basic epidemic behavior. 

```{r, tidy=FALSE}

standard = function(x)
{
    (x-mean(x))/sd(x)  
}
# Fixed Co-variates
X = matrix(1)
# Time Varying co-variates
daysSince2000 = as.numeric(as.Date(fluDat.sub$Date) - as.Date("2000-01-01"))
trigBasis = daysSince2000/365*2*pi
Z = cbind(standard(daysSince2000),            # linear term for time
          sin(trigBasis),               # trig basis function 1
          cos(trigBasis),               # trig basis function 2
          sin(2*trigBasis),
          cos(2*trigBasis),
          sin(0.5*trigBasis),
          cos(0.5*trigBasis)
  )

# There is only one spatial location here, so we don't need to do anything 
# else to Z. If there were, and the time varying co-variates were the same 
# in each location, we could just duplicate Z row-wise:
# Z = Z[rep(1:nrow(Z), each = nrow(X)),]
```

Let's look at our infection process basis (minus the intercept).

```{r, tidy=FALSE, echo=TRUE}

par(mfrow = c(2,2))
for (i in 1:4)
{
  par(xaxt = "n")
  plot(Z[,i], type = "l", main = paste("Time Varying Co-variate", i),
       xlab = "Year", ylab = "Value")
  xAxis()
}
```



2b: Immunity Loss Process
---------------------------
The same basis used for the infection process may serve well for the reintroduction of recovered individuals to the susceptible category. In addition, we may wish to evaluate to what degree individuals were more susceptible to the H1N1 pandemic in 2009. We therefore introduce a time varying H1N1 term, which is zero before April 2009, jumps to one on April first, and then decays. 


```{r, tidy=FALSE, echo=TRUE}
h1n1_term = (as.Date(fluDat.sub$Date) >= as.Date("2009-04-01"))
h1n1_idx = which(h1n1_term)
h1n1_term[h1n1_idx] = h1n1_term[h1n1_idx]/sqrt(seq(1,length(h1n1_idx)))

X_p_rs = cbind(as.numeric(X), Z, h1n1_term)

par(xaxt="n")
plot(h1n1_term, type = "l", lwd = 2, 
     main = "H1N1 Susceptibility Co-variate",
     xlab = "Year", ylab= "Value")
xAxis()
```


2c: Parameter Starting Values
------------------------------


```{r, tidy=FALSE}

I_star = matrix(round(FluCountEst), ncol = 1)
N = matrix(TotalPop, nrow = nrow(fluDat.sub), ncol = 1)

# The following parameters determine p_ei and p_ir
# if the offset is all ones, p_ei = 1-exp(-gamma_ei) etc.
gamma_ei = 2.3
gamma_ir = 2.3

beta = rep(0, ncol(X) + ncol(Z))
beta[1] = 2.5

beta_p_rs = rep(0, ncol(X_p_rs))
beta_p_rs[1] = -2.5


# Dummy values for spatial components, as this is a non spatial model.
rho = -1 # spatial dependence parameter
DM = c(0)
scaleDistMode = 0


# Starting tuning parameters for MCMC sampling
mcmcTuningParams = c(1, # S_star
                     1, # E_star
                     1,  # R_star
                     1,  # S_0
                     1,  # I_0
                     0.82,  # beta
                     0.2,  # beta_p_rs
                     0.0, # rho
                     0.01, # gamma_ei
                     0.01  # gamma_ir
                     )  


# Prior parameters. 
priorAlpha_gammaEI = 230;
priorBeta_gammaEI = 100;
priorAlpha_gammaIR = 230;
priorBeta_gammaIR = 100;
betaPrsPriorPrecision = 0.1
betaPriorPrecision = 0.1

# reinfection mode is 1, because this is a general SEIRS model. 
reinfectionMode = 1
# steadyStateConstraintPrecision is a loose constraint on net flows
# between compartments. Setting it to a negative value eliminates
# the constraint, but it can help with identifiability in some cases. 
steadyStateConstraintPrecision = 0.1

# iterationStride determines the delay between saving samples to the output file
iterationStride = 10000
outFileName = "./chain_output.txt"

# Make a crude guess as to the true compartments:
# S_star, E_star, R_star, and thus S,E,I and R
proposal = generateCompartmentProposal(I_star, N, 
                                       I0 = 100000, 
                                       p_ir = 0.9, 
                                       p_rs = 0.05)

# Turn off verbose and debug output.
verbose = FALSE
debug = FALSE


# Get object dimensions. This will be done automatically in the future
compMatDim = dim(I_star)
xDim = dim(X)
zDim = dim(Z)
xPrsDim = dim(X_p_rs)

# Create the offset. All time points are evenly spaced, so this is just a vector of ones. 
offset = rep(1, nrow(I_star))
```

Part 3: Model Fitting
=======================

We're ready to build the model. It would be preferable to build more than one model to assess chain convergence, but for illustrative purposes this tutorial takes the simpler approach of using one chain. 


```{r, tidy=FALSE}
# Create model object. This giant constructor is going to be replaced with something friendlier 
# in the future. 
SEIRmodel = spatialSEIRModel(compMatDim,
                      xDim,
                      zDim,
                      xPrsDim,
                      proposal$S0,
                      proposal$E0,
                      proposal$I0,
                      proposal$R0,
                      proposal$S_star,
                      proposal$E_star,
                      proposal$I_star,
                      proposal$R_star,
                      offset,
                      X,
                      Z,
                      X_p_rs,
                      DM,
                      rho,
                      priorAlpha_gammaEI,
                      priorBeta_gammaEI,
                      priorAlpha_gammaIR,
                      priorBeta_gammaIR,
                      beta,
                      betaPriorPrecision,
                      beta_p_rs,
                      betaPrsPriorPrecision,
                      gamma_ei,
                      gamma_ir,
                      N,
                      outFileName,
                      iterationStride,
                      steadyStateConstraintPrecision,
                      verbose,
                      debug,
                      mcmcTuningParams,
                      reinfectionMode,
                      scaleDistMode)

# Sample regression parameters individually.
SEIRmodel$parameterSamplingMode=1

```

Before we do any sampling, let's tell spatialSEIR to keep track of the sampling behavior of the last time point so that we can do prediction later. 

```{r, tidy=FALSE}

SEIRmodel$setTrace(0, (nrow(SEIRmodel$S) - 1))

```

3a: Tuning Parameters
-------------------------

Now we're ready to start sampling. On the other hand, we're not particularly confident that we've used reasonable tuning parameters. Let's add a function to help choose these. 


```{r tidy=FALSE}
burnInSamplingParams = function(modelObject,
                               numBatches=500, 
                               batchSize=20, 
                               targetAcceptanceRatio=0.05,
                               tolerance=0.0025,
                               proportionChange = 0.1
                              )
{
    for (batch in 1:numBatches)
    {
        modelObject$simulate(batchSize)
        modelObject$updateSamplingParameters(targetAcceptanceRatio, tolerance, proportionChange)
    }
}

```

Before we adjust these, however, we may be interested whether using the OpenCL capabilities is any faster for this particular data size. Let's take a look at what devices we have available:

```{r tidy=FALSE}

SEIRmodel$printOCLSummary()

```

There is only one device on this laptop, the (0,0) device. It is already, therefore, the current device. If there were more present, they could be selected with the following syntax:

```{r tidy=FALSE}
SEIRmodel$setDevice(0,0)
```

Let's take a look at how long 1000 iterations take in CPU mode (the default):

```{r tidy=FALSE}
system.time(SEIRmodel$simulate(1000))
```
Note that the "elapsed" number is the important one as far as we're concerned. 
Now let's tell spatialSEIR that we want to use OpenCL and try again.

```{r tidy=FALSE}
SEIRmodel$oclPreferences = SEIRmodel$oclPreferences + 1
system.time(SEIRmodel$simulate(1000))
```

Looks like CPU mode wins handily for this small data set, so let's switch our preferences back. 

```{r tidy=FALSE}
SEIRmodel$oclPreferences = SEIRmodel$oclPreferences-1
```

Now we can worry about our tuning parameters. Let's run the sampler for a bit, and then look at our acceptance rates. 

```{r tidy=FALSE}

# Adjust the sampling parameters

burnInSamplingParams(SEIRmodel)

#Check the current acceptance rates. First simulate, as the counters are cleared each time the parameters 
# are updated:
SEIRmodel$simulate(1000)
SEIRmodel$printAcceptanceRates()
SEIRmodel$printSamplingParameters()

```

These look reasonable, so let's launch the main simulation. As currently implemented, the MCMC samplers encounter distressingly high autocorrelation. This is an active target of optimization, but in the meantime we need to be prepared to run inconveniently many samples. A bit of timekeeping code helps keep track of how long this might take. 


```{r tidy = FALSE}
# Switch to block sampling for parameters
SEIRmodel$parameterSamplingMode=3

tm = 0
for (i in 1:100)
{
  tm = tm + system.time(SEIRmodel$simulate(100000))[3]
  SEIRmodel$updateSamplingParameters(0.05, 
                                     0.0025, 
                                     0.05)
  cat(paste(i, "of 100,", 
            round((tm/(i)*(100-i)/60), 2), 
            " minutes remaining.\n"))
}

```

3b: Initial Results
-----------------------------

Let's take a look at the estimated epidemic. 

```{r tidy = FALSE}
  S_frac = SEIRmodel$S/N
  E_frac = SEIRmodel$E/N
  I_frac = SEIRmodel$I/N
  R_frac = SEIRmodel$R/N

  maxProp = max(c(S_frac, E_frac, I_frac, R_frac))
  par(xaxt="n")
  plot.new()
  layout(matrix(c(1,2), nrow = 1),
        widths = c(8,2), heights = c(4,4))
  par(bty = "o")
  plot(0,0, xlim = c(0, nrow(S_frac)), 
       ylim = c(0, min(maxProp*1.1, 1.0)), type = "n",
       main = "Estimated Epidemic Fractions",
       xlab = "Year", ylab = "Proportion")
  xAxis()

  lines(S_frac, col = "blue", lty = 1, lwd = 1.5)
  lines(E_frac, col = "red", lty = 2, lwd = 1.5)
  lines(I_frac, col = "green", lty = 3, lwd = 1.5)
  lines(R_frac, col = "black", lty = 4, lwd = 1.5)
  par(bty = "n")
  par(xaxt="n"); par(yaxt="n")
  plot(c(0,10), c(0,10), type = "n",xlab="",ylab="")
  par(xpd=TRUE)
  legend(x=-55,y=10,legend=c("Susceptible", "Exposed", "Infectious", "Removed"),
         lty = 1:4, col = c("blue", "red", "green", "black"), lwd = 1.5, cex = 0.7)
par(xpd=FALSE)
```

We could also use color to indicate epidemic intensity (more useful with multiple locations).

```{r tidy=FALSE, fig.width=6,fig.height=4}
  {
  plotCompartment(SEIRmodel$I/N)
  }
```

Let's examine the parameter chains. Again, these should be started from multiple points in the 
parameter space and assessed for convergence in a real analysis. 

```{r tidy=FALSE}
chainDat = read.csv("./chain_output.txt")
names(chainDat)

plotChain = function(chain, main)
{
  chainIdx = floor(length(chain) - length(chain)/2):length(chain)
  plot(chainDat$Iteration[chainIdx],chain[chainIdx],
       main = main, xlab = "Iteration", ylab = "Value", type = "l")
  abline(h = mean(chain[chainIdx]), lty = 2, col = "grey")
}
```

```{r, tidy=FALSE,echo=TRUE, fig.width=4, fig.height=12}
par(mfrow = c(5,2))
plotChain(chainDat$BetaP_SE_0, "Exposure Intercept")
plotChain(chainDat$BetaP_RS_0, "Reinfection Intercept")

plotChain(chainDat$BetaP_SE_1, "Exposure Linear Time Component")
plotChain(chainDat$BetaP_RS_1, "Reinfection Linear Component")

plotChain(chainDat$BetaP_SE_2, "Exposure Sin Component")
plotChain(chainDat$BetaP_RS_2, "Reinfection Sin Component")

plotChain(chainDat$BetaP_SE_3, "Exposure Cos Component")
plotChain(chainDat$BetaP_RS_3, "Reinfection Cos Component")

plot.new()
plotChain(chainDat$BetaP_RS_4, "Reinfection H1N1 Component")
```
```{r, tidy=FALSE, echo=TRUE, fig.width=4, fig.height=8}
par(mfrow = c(2,1))
plotChain(1-exp(-chainDat$gamma_ei), "E to I Transition Probability")
plotChain(1-exp(-chainDat$gamma_ir), "I to R Transition Probability")
```


Clearly, autocorrelation is a major issue. Simulations do, however, indicate that the sampler will converge eventually to a stable posterior encompasing the true values of the parameters. 


3c: Basic Reproductive Number
--------------------------------
An additional task we will often want to do is to calculate the basic reproductive number over time. This is partially implemented in the spatialSEIR library, and can be completed easily in R. 


```{r, tidy =FALSE}
  getR0 = function(t)
  {
    max(eigen(SEIRmodel$getGenerationMatrix(t))$values)  
  }
  
  R0_vec = sapply(1:(nrow(SEIRmodel$I)-1), getR0)

  par(xaxt = "n")
  plot(R0_vec, type = "l", xlab = "Year",ylab = "R0",
       main = "Estimated Basic Reproductive Number")
  abline(h=seq(0, 2, 0.1), lty=2, col="lightgrey")
  abline(h = 1.0, col = "blue", lwd = 1.5, lty = 2)
  lines(R0_vec, lwd = 1.5)
  xAxis()
```

This could be easily done periodically during sampling to obtain posterior R0 distribution estimates. 

3d: Prediction
----------------

Prediction is not built into the spatialSEIR library as of yet, but we can still make predictions manually. Let's try to predict the 2011 epidemic season. The following is just a crude estimate based on the parameter means. It would be preferable to take a sample of 
the parameter values and predict the epidemic for each. 


```{r tidy=FALSE}

  predictEpidemic = function(beta.pred, 
                             beta.prs.pred, 
                             X.pred,
                             X.prs.pred,
                             gamma.ei,
                             gamma.ir,
                             S0,
                             E0,
                             I0,
                             R0)
  {
      N = (S0+E0+I0+R0)
      p_se_components = exp(X.pred %*% beta.pred)
      p_se = matrix(0, ncol = 1, nrow = nrow(X.pred))
      p_rs = exp(X.prs.pred %*% beta.prs.pred)
      p_ei = 1-exp(-gamma.ei)
      p_ir = 1-exp(-gamma.ir)
      S_star = matrix(0, ncol=1,nrow = nrow(X.pred))
      E_star = matrix(0, ncol=1,nrow = nrow(X.pred))
      I_star = matrix(0, ncol=1,nrow = nrow(X.pred))
      R_star = matrix(0, ncol=1,nrow = nrow(X.pred))
      S = matrix(0, ncol=1,nrow = nrow(X.pred))
      E = matrix(0, ncol=1,nrow = nrow(X.pred))
      I = matrix(0, ncol=1,nrow = nrow(X.pred))
      R = matrix(0, ncol=1,nrow = nrow(X.pred))
      S[1,] = S0
      E[1,] = E0
      I[1,] = I0
      R[1,] = R0
      S_star[1,] = rbinom(1, R0, p_rs[1])
      p_se[1,] = 1-exp(-(I0/N*p_se_components[1]))
      E_star[1,] = rbinom(1, S0, p_se[1,])
      I_star[1,] = rbinom(1, E0, p_ei)
      R_star[1,] = rbinom(1, I0, p_ir)
      
      for (i in 2:nrow(S))
      {
      
        S[i,] = S[i-1,] + S_star[i-1,] - E_star[i-1,]
        E[i,] = E[i-1,] + E_star[i-1,] - I_star[i-1,]
        I[i,] = I[i-1,] + I_star[i-1,] - R_star[i-1,]
        R[i,] = R[i-1,] + R_star[i-1,] - S_star[i-1,]
        
        p_se[i,] = 1-exp(-(I[i,]/N*p_se_components[i]))
        S_star[i,] = rbinom(1, R[i,], p_rs[i])
        E_star[i,] = rbinom(1, S[i,], p_se[i,])
        I_star[i,] = rbinom(1, E[i,], p_ei)
        R_star[i,] = rbinom(1, I[i,], p_ir)
      }
      return(list(S=S,E=E,I=I,R=R,
                  S_star=S_star,E_star=E_star,
                  I_star=I_star,R_star=R_star,
                  p_se=p_se,
                  p_rs=p_rs))
  }

  dat.pred = ((as.Date(fluDat$Date) >= as.Date("2011-01-01")) & 
              (as.Date(fluDat$Date) < as.Date("2012-01-01")))
  fluDat.pred = fluDat[dat.pred,]

  daysSince2000.pred = as.numeric(as.Date(fluDat.pred$Date) - as.Date("2000-01-01"))
  daysSince2000.pred = (daysSince2000.pred - mean(daysSince2000))/sd(daysSince2000)
  trigBasis.pred = daysSince2000.pred/365*2*pi
  Z.pred = cbind(daysSince2000.pred,            # linear term for time
            sin(trigBasis.pred),               # trig basis function 1
            cos(trigBasis.pred))               # trig basis function 2
  X.pred = cbind(1,Z.pred)
  # re-infection prediction co-variates
  h1n1_term.pred = (as.Date(fluDat.pred$Date) >= as.Date("2009-04-01"))
  h1n1_idx.pred = which(h1n1_term.pred)
  h1n1_term.pred[h1n1_idx.pred] = h1n1_term.pred[h1n1_idx.pred]/sqrt(seq(1,length(h1n1_idx.pred)))

  X.prs.pred = cbind(1, Z.pred, h1n1_term.pred)


  predict.i = function(i)
  {
    dataRow = chainDat[i,]
    beta = c(dataRow$BetaP_SE_0,
             dataRow$BetaP_SE_1,
             dataRow$BetaP_SE_2,
             dataRow$BetaP_SE_3)
    beta.prs = c(dataRow$BetaP_RS_0,
                 dataRow$BetaP_RS_1,
                 dataRow$BetaP_RS_2,
                 dataRow$BetaP_RS_3,
                 dataRow$BetaP_RS_4)
    return(predictEpidemic(beta, 
                           beta.prs, 
                           X.pred,
                           X.prs.pred,
                           dataRow$gamma_ei,
                           dataRow$gamma_ir,
                           dataRow$S_0_207 + dataRow$S_star_0_207 - dataRow$E_star_0_207,
                           dataRow$E_0_207 + dataRow$E_star_0_207 - dataRow$I_star_0_207,
                           dataRow$I_0_207 + dataRow$I_star_0_207 - dataRow$R_star_0_207,
                           dataRow$R_0_207 + dataRow$R_star_0_207 - dataRow$S_star_0_207))
                                             # These values are made available by setTrace, 
    }
  
  preds = lapply((nrow(chainDat) - floor(nrow(chainDat)/2)):
                   nrow(chainDat), predict.i)
  FluCountEst.pred = (fluDat.pred$HHS.Region.5..IL..IN..MI..MN..OH..WI./100000)*TotalPop
```

There's a lot going on with the predictions - let's focus on the infection size for simplicity. 

```{r, tidy=FALSE}
par(xaxt="n") # Suppress plotting the X axis
# Plot the data with room for the prediction
plot(SEIRmodel$I_star, type = "l", 
     main = "ILI Per 100k Physician Visits\n HHS Region 5", 
     xlab = "Year", ylab = "#Visits/100k",
     xlim = c(0, length(region5) + nrow(X.pred)))
# Add horizontal reference lines
abline(h=seq(0, 10000000,500000), lty = 2, col = "lightgrey")
# Prepare expanded x axis plotting function
lines(region5, col = "blue", lwd = 1.5)
years.pred = as.numeric(format(c(as.Date(fluDat.sub$Date), as.Date(fluDat.pred$Date)), "%Y"))
yearIdx.pred = which(!duplicated(years.pred))
xAxis.pred = function()
{
  par(xaxt="s")  
  axis(side=1,at=yearIdx.pred, labels = years.pred[yearIdx.pred])
}
xAxis.pred()
# Prepare to plot predictions
pred.color = rgb(0.1,0.1,0.1,0.1)
pred.idx = c((nrow(SEIRmodel$S)+1):(nrow(SEIRmodel$S)+nrow(X.pred)))
# Plot MCMC predictions
for (pred in preds)
{
  lines(pred.idx,pred$I_star, 
        lty=2, col = pred.color)
}
# Plot true, unobserved data in red. 
lines(c((nrow(SEIRmodel$S)+1):(nrow(SEIRmodel$S)+nrow(X.pred))),
      FluCountEst.pred, lty=3, col = "red", lwd = 2)

```

It looks like the prediction does fairly well until the next peak, but then doesn't decrease quickly enough. This may have to do with the relatively simple basis used to drive the exposure process. Clearly theres a lot of work to do!

That wraps it up for now - the next tutorial will describe the spatial features of spatialSEIR. The library is under active development, so much of the previous work which was done with R code will be provided by the model object. In addition, sampling efficiency is an important goal which is being explored, so the library will hopefully require less sampling time in the near future. 


